---
title: "Model_Validation"
output: html_document
date: "2025-06-03"
---


This document contains the code to reproduce all supplementary Figures of our paper as well as some additional figures to validate our approach of estimating $K_D$ estimates from titration binding curves

```{r}
#Switch to False if you wish to perform fitting of all artificial data from scratch
#Keep at True if you wish to skip the computationally expensive simulations and read in previously simulated results
load_results = T
```

```{r}
library(latex2exp)
library(deSolve)
library(clinPK)
library(ggplot2)
library(dplyr)
library(minpack.lm)
library(tidyr)
library(ggbeeswarm)
library(Hmisc)
library(scales)
library(pheatmap)
source('Functions.R')
```

```{r}
#Set seed in order to ensure reproducability
set.seed(20250602)
```


---------
Figure 3a
---------

Provides an overview of the approach for estimating $K_D$
```{r}
singlemodelFit(MFI_df = Hap1WT_df, k2_guess = .2, SF_guess = 100, celltype = 'Hap1')
```


----------------------
Supplemental Figure 3a
----------------------

```{r}
#Generate dose-response data for L_init = 2
range.dose = seq(0,5,.01)
doseresp1 = matrix(nrow=length(range.dose), ncol=4)
j=1
pars1 = c(k1=1.0, k2=0.1)
for (i in range.dose) {
  inistate = c(A=i, L=2, C=0)
  out = ode(y=inistate, times=seq(0,100,1), func=affinity.ODE.1ligand, parms=pars1)
  doseresp1[j,] = c(i, out[nrow(out),4], out[nrow(out),4]/(out[nrow(out),3]+out[nrow(out),4]), out[nrow(out),2])
  #doseresp1[j,] = c(i, out[3,4] + out[3,6])
  j=j+1
}

#Generate dose-response data for L_init = 5
doseresp2 = matrix(nrow=length(range.dose), ncol=4)
j=1
pars2 = c(k1=1.0, k2=0.1)
for (i in range.dose) {
  inistate = c(A=i, L=0.5, C=0)
  out = ode(y=inistate, times=seq(0,100,1), func=affinity.ODE.1ligand, parms=pars2)
  doseresp2[j,] = c(i, out[nrow(out),4], out[nrow(out),4]/(out[nrow(out),3]+out[nrow(out),4]), out[nrow(out),2])
  #doseresp2[j,] = c(i, out[3,4] + out[3,6])
  j=j+1
}

doseresp1 = as.data.frame(doseresp1)
doseresp2 = as.data.frame(doseresp2)
names(doseresp1) = names(doseresp2) = c('A_ini', 'C_eq', 'Frac_boundL', 'A_eq')

#identify the inflection points of the curves
CI_AiniWT = doseresp1$A_ini[which(abs(doseresp1$C_eq - 1) == min(abs(doseresp1$C_eq - 1)))]
CI_AiniKO = doseresp2$A_ini[which(abs(doseresp2$C_eq - .25) == min(abs(doseresp2$C_eq - .25)))]
CI_AeqWT = doseresp1$A_eq[which(abs(doseresp1$C_eq - 1) == min(abs(doseresp1$C_eq - 1)))]
CI_AeqKO = doseresp2$A_eq[which(abs(doseresp2$C_eq - .25) == min(abs(doseresp2$C_eq - .25)))]

#Add a group column for coloring
doseresp1$`initial [L]` = "2"
doseresp2$`initial [L]` = ".5"

#Combine both into one data frame
doseresp_combined = rbind(doseresp1, doseresp2)

ggplot(doseresp_combined, aes(x = A_ini, y = C_eq, color = `initial [L]`)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("2" = "#486a8b", ".5" = "#a55d5d")) +
  labs(x = TeX('$\\[A_{init}\\]$ (a.u.)'), y = TeX('$\\[C_{eq}\\]$ (a.u.)')) +
  ylim(0, 2) +
  geom_hline(yintercept = c(1), linetype = "dashed", color = "#486a8b", linewidth = 1) +
  geom_vline(xintercept = CI_AiniWT, linetype = "dashed", color = "#486a8b", linewidth = 1) +
  geom_hline(yintercept = c(0.25), linetype = "dashed", color = "#a55d5d", linewidth = 1) +
  geom_vline(xintercept = CI_AiniKO, linetype = "dashed", color = "#a55d5d", linewidth = 1) +
  geom_point(aes(x = CI_AiniKO, y = 0.25), color = "#a55d5d", size = 4) + #Add large dot
  annotate("text", x = CI_AiniKO + .1, y = 0.25 - .08, label = TeX(sprintf("$EC_{50}$ = %.2f", CI_AiniKO)), hjust = 0, size = 4, fontface = 'bold') + #Add text annotation to the right
  geom_point(aes(x = CI_AiniWT, y = 1), color = "#486a8b", size = 4) + #Add large dot
  annotate("text", x = CI_AiniWT + .1, y = 1 - .08, label = TeX(sprintf("$EC_{50}$ = %.2f", CI_AiniWT)), hjust = 0, size = 4) + #Add text annotation to the right
  theme_classic()

ggsave('../Figures/Supporting_figures/EC50pred_iniCurve.png', width = 5, height = 3, dpi = 600)

ggplot(doseresp_combined, aes(x = A_eq, y = C_eq, color = `initial [L]`)) +
  geom_line(linewidth = 1.5) +
  scale_color_manual(values = c("2" = "#486a8b", ".5" = "#a55d5d")) +
  labs(x = TeX('$\\[A_{init}\\]$ (a.u.)'), y = TeX('$\\[C_{eq}\\]$ (a.u.)')) +
  ylim(0, 2) +
  geom_hline(yintercept = c(1), linetype = "dashed", color = "#486a8b", linewidth = 1) +
  geom_vline(xintercept = 0.1, linetype = "solid", color = "#486a8b", linewidth = 1) +
  geom_hline(yintercept = c(0.25), linetype = "dashed", color = "#a55d5d", linewidth = 1) +
  geom_vline(xintercept = .1, linetype = "dashed", color = "#a55d5d", linewidth = 1.2) +
  geom_point(aes(x = CI_AeqKO, y = 0.25), color = "#a55d5d", size = 4) + #Add large dot
  annotate("text", x = CI_AeqKO + .1, y = 0.25 - .08, label = TeX(sprintf("$EC_{50}$ = %.2f", CI_AeqKO)), hjust = 0, size = 4, fontface = 'bold') + #Add text annotation to the right
  geom_point(aes(x = CI_AeqWT, y = 1), color = "#486a8b", size = 4) + #Add large dot
  annotate("text", x = CI_AeqWT + .1, y = 1 - .08, label = TeX(sprintf("$EC_{50}$ = %.2f", CI_AeqWT)), hjust = 0, size = 4) + #Add text annotation to the right
  theme_classic()

ggsave('../Figures/Supporting_figures/EC50pred_eqCurve.png', width = 5, height = 3, dpi = 600)
```


------------------------
Supplemental Figure 3b-c
------------------------

```{r}
check_maxbias(SF = 350, varconstant_within = 0, varconstant_between =0, k1 = 1, k2 = 100, L_init = 40, nrep = 1)
check_maxbias(SF = 350, varconstant_within = 0, varconstant_between =0, k1 = 1, k2 = 1, L_init = 40, nrep = 1)
check_maxbias(SF = 350, varconstant_within = 0, varconstant_between =0, k1 = 1, k2 = .01, L_init = 40, nrep = 1)

check_maxbias(SF = 350, varconstant_within = 0, varconstant_between =0, k1 = 1, k2 = 100, L_init = 40, nrep = 1, top5 = T)
check_maxbias(SF = 350, varconstant_within = 0, varconstant_between =0, k1 = 1, k2 = 1, L_init = 40, nrep = 1, top5 = T)
check_maxbias(SF = 350, varconstant_within = 0, varconstant_between =0, k1 = 1, k2 = .01, L_init = 40, nrep = 1, top5 = T)
```


----------------------
Supplemental Figure 3d
----------------------

```{r}
k2_range = exp(seq(log(0.01), log(500), length.out = 500))
k2_range = 10^seq(log10(0.01), log10(500), length.out = 500)

deviance = deviance5 = numeric(length(k2_range))

for(i in 1:length(k2_range)){
  deviance[i] = check_maxbias(k1 = 1, k2 = k2_range[i], top5 = F, plot = F)
  deviance5[i] = check_maxbias(k1 = 1, k2 = k2_range[i], top5 = T, plot = F)
}

df = data.frame(
  k2 = k2_range,
  `Full range` = deviance,
  `Highest 5` = deviance5,
  check.names = F
)

df_long = pivot_longer(df, cols = c(`Full range`, `Highest 5`), names_to = "type", values_to = "value")

#Create plot
full_vs_high5 = ggplot(df_long, aes(x = k2, y = value, color = type)) +
  geom_line(linewidth = 1.1) +
  geom_hline(aes(yintercept = 14000, color = "True max"), 
             linetype = "dashed", linewidth = 1.1) +
  scale_x_log10(labels = function(x) format(x, scientific = FALSE)) +
  scale_color_manual(
    name = "Fitted conc. range",
    values = c("Full range" = "#486a8b", 
               "Highest 5"  = "#a55d5d", 
               "True max"  = "black")
  ) +
  coord_cartesian(ylim = c(13500, 15500)) +
  labs(
    x = TeX('$K_D$ (a.u.)'), 
    y = TeX('$MFI_{max}$')
  ) +
  theme_classic()

print(full_vs_high5)

ggsave('../Figures/Supporting_figures/full_vs_high5.1.png', full_vs_high5, width = 5, height = 3)

```


--------------------------
Generating artificial data
--------------------------

```{r}
#Baseline artificial data
art0WT_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 90, L_init = 40)
art0KO_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 180, L_init = 10)

#Initial concentrations of L are the same
art1WT_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 90, L_init = 40)
art1KO_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 180, L_init = 40)

#Reduced scaling factor
art2WT_list = create_art_df_list(n = 10, SF = 150, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 90, L_init = 40)
art2KO_list = create_art_df_list(n = 10, SF = 150, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 180, L_init = 10)

#Increased variance
art3WT_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .02, varconstant_between = .2, k1 = 3, k2 = 90, L_init = 40)
art3KO_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .02, varconstant_between = .2, k1 = 3, k2 = 180, L_init = 10)

#Different KDs
art4WT_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 160, L_init = 40)
art4KO_list = create_art_df_list(n = 10, SF = 350, varconstant_within = .01, varconstant_between = .1, k1 = 3, k2 = 40, L_init = 10)

#Estimating paramater values from artificial data

if(!load_results){
  out_art0 = oneligand_fit_list(art0WT_list, art0KO_list, label_WT = "art0_WT", label_KO = "art0_KO", celltype = 'art0')
  out_art1 = oneligand_fit_list(art1WT_list, art1KO_list, label_WT = "art1_WT", label_KO = "art1_KO", celltype = 'art1')
  out_art2 = oneligand_fit_list(art2WT_list, art2KO_list, label_WT = "art2_WT", label_KO = "art2_KO", celltype = 'art2')
  out_art3 = oneligand_fit_list(art3WT_list, art3KO_list, label_WT = "art3_WT", label_KO = "art3_KO", celltype = 'art3')
  out_art4 = oneligand_fit_list(art4WT_list, art4KO_list, label_WT = "art4_WT", label_KO = "art4_KO", celltype = 'art4')
  
  saveRDS(out_art0, '../Results/out_art0.rds')
  saveRDS(out_art1, '../Results/out_art1.rds')
  saveRDS(out_art2, '../Results/out_art2.rds')
  saveRDS(out_art3, '../Results/out_art3.rds')
  saveRDS(out_art4, '../Results/out_art4.rds')
}else{
  out_art0 = readRDS('../Results/out_art0.rds')
  out_art1 = readRDS('../Results/out_art1.rds')
  out_art2 = readRDS('../Results/out_art2.rds')
  out_art3 = readRDS('../Results/out_art3.rds')
  out_art4 = readRDS('../Results/out_art4.rds')
}
```


----------------------
Supplemental Figure 3e
----------------------

Show that artificial data mimics experimental data
```{r}
#Hap1WT_df = read_MFI_data('../Data/Experimental_Data/Antibody_Titrations.xlsx', sheet = 'Hap1 WT', molar_weight_AB = 146000)

ggplot(Hap1WT_df, aes(x = conc, y = MFI, shape = replicate)) +
  geom_point(size = 2, col = 'grey') + 
  labs(title = 'Hap1 WT') +
  scale_x_log10() +
  theme_classic() + 
  theme(panel.grid = element_blank())

ggsave('../Figures/Supporting_figures/Hap1WT_datareps.png', width = 4, height = 3, dpi = 500)

ggplot(art0WT_list[[10]], aes(x = conc, y = MFI, shape = replicate)) +
  geom_point(size = 2, col = 'grey') + 
  labs(title = 'set0 WT') +
  scale_x_log10() +
  theme_classic() + 
  theme(panel.grid = element_blank())

ggsave('../Figures/Supporting_figures/art0.1WT_datareps.png', width = 4, height = 3, dpi = 500)
```


------------------------
Supplemental Figure 3g-i
------------------------

```{r, warning=F}
#Prepare data for plotting

dfs = list(out_art0, out_art1, out_art2, out_art3, out_art4)
names(dfs) = paste0("df", 0:4)

#Combine into long format
long_df = bind_rows(dfs, .id = "source")

#Prepare long format for k2
k2_long = long_df %>%
  filter(source %in% c('df0', 'df1', 'df2', 'df3')) %>%
  select(source, k2_WT, k2_KO) %>%
  pivot_longer(cols = c(k2_WT, k2_KO),
               names_to = "genotype",
               names_prefix = "k2_",
               values_to = "k2") %>%
  mutate(pair = paste0("pair", source))

#Same for max
max_long = long_df %>%
  filter(source %in% c('df0','df3', 'df4')) %>%
  select(source, WT_max, KO_max) %>%
  pivot_longer(
    cols = c(WT_max, KO_max),
    names_to = "genotype",
    values_to = "max"
  ) %>%
  mutate(
    genotype = case_when(
      grepl("^WT", genotype) ~ "WT",
      grepl("^KO", genotype) ~ "KO"
    ),
    pair = paste0("pair", source)
  )

plot_single_metric(long_df[long_df$source != 'df2',], "SF", ref_line = 350, log_scale = T)
plot_single_metric(long_df, "AIC", log_scale = F)

plot_paired_metric(k2_long, "k2", ylab = TeX('$K_D$ estimates'), ref_lineWT = 30, ref_lineKO = 60)  #No ref line
plot_paired_metric(max_long, "max", ylab = TeX('$MFI_{max}$ estimates'), ref_lineWT = 350 * 40, ref_lineKO = 350 * 10)
```


--------------------------------------------------
Extra check to ensure fitting algorithm works well
--------------------------------------------------

See if fitting algorithm works properly by examining RMSE landscape over a range of k2 and SF values
```{r}
MFI_df = Hap1WT_df
k2_guess = .2
SF_guess = 100
celltype = 'Hap1'

MFI_max = determine_max(MFI_df, label = celltype, returnplot = F)

k2s = 0:99
SFs = 0:99

RMSEs = matrix(nrow = length(k2s), ncol = length(SFs))

for (i in 1:length(k2s)) {
  for (j in 1:length(SFs)) {
    RMSEs[i, j] = tryCatch({
      resids = residual_function1(
        params = c(k2 = k2s[i], SF = SFs[j]),
        A_init = MFI_df$conc,
        MFI_obs = MFI_df$MFI,
        k1 = 1,
        MFImax = MFI_max
      )
      round(sqrt(mean(resids^2)), digits = 1)
    }, error = function(e) {
      NA
    })
  }
}


pheatmap(RMSEs, cluster_rows = F, cluster_cols = F, labels_row = 'k2', labels_col = 'SF')

min_index = which(RMSEs[,-1] == min(RMSEs[,-1]), arr.ind = T) + c(0,1)
original_min = RMSEs[min_index]
RMSEs[min_index] = 1500


pheatmap(RMSEs, cluster_rows = F, cluster_cols = F, labels_row = 'k2', labels_col = 'SF')
```


